---
title: "Plotting Multiple Sequence Alignment Chord Diagrams with 'ggchord'"
author: "DangJem"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggchord Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 7
)
```

## 1. Introduction

`ggchord` is a `ggplot2`-based R package for visualizing multi-sequence alignment results as intuitive chord diagrams. It uses a layered grammar of graphics: `ggchord()` validates the data and sets global style, and the `geom_*` layers (`geom_seq()`, `geom_ribbon()`, `geom_gene()`, `geom_axis()`) each manage their own layout parameters and rendering. The layout is computed when the plot is printed, so parameters can be spread across the layers just like in `ggplot2`.

This tutorial walks through the main functions and parameters using the built-in example data.

## 2. Installation and Loading

Install the development version from GitHub (or install a local tarball), then load the package:

```{r install, eval=FALSE}
# Install from GitHub
# devtools::install_github("DangJem/ggchord")

# Or install from a local source tarball
# devtools::install_local("path/to/ggchord_0.5.0.tar.gz")

# Load the package
library(ggchord)
library(dplyr)  # For data processing
```

```{r load, echo=FALSE}
library(ggchord)
library(dplyr)
```

## 3. Data Preparation

`ggchord` requires three types of data (only sequence data is mandatory):

- Sequence data (`seq_data`): Contains sequence IDs and lengths
- Alignment data (`ribbon_data`): pairwise alignment results for drawing inter-sequence ribbons
- Gene annotation data (`gene_data`): Contains gene positions and annotation information

This tutorial uses the built-in example data:

```{r load-data}
# Load built-in example data
data(seq_data_example)    # Sequence data
data(ribbon_data_example) # Alignment data
data(gene_data_example)   # Gene annotation data

# Inspect data structure
head(seq_data_example)
head(ribbon_data_example)
head(gene_data_example)
```

## 4. Basic Usage

### 4.1 Using Only Sequence Data

Sequence data is fundamental for drawing chord diagrams, requiring at minimum `seq_id` (sequence ID) and `length` (sequence length):

```{r part1-1}
# Basic chord diagram: sequences only
p1 <- ggchord(seq_data = seq_data_example) + geom_seq()
print(p1)
```

Customize sequence order, orientation, curvature, and colors via `geom_seq()`:

```{r part1-2}
p2 <- ggchord(seq_data = seq_data_example) +
  geom_seq(
    seq_order = c("MT118296.1", "OR222515.1", "MT108731.1", "OQ646790.1"), # Sequence order
    seq_orientation = c(1, -1, 1, -1),  # Direction (1 = forward, -1 = reverse)
    seq_curvature = c(0, 2, -2, 6),     # Curvature (0 = straight line, 1 = standard arc)
    seq_colors = c("steelblue", "orange", "pink", "yellow")  # Sequence colors
  )
print(p2)
```

### 4.2 Adding Sequence Alignment Data

Alignment data (`ribbon_data`) visualizes sequence similarity, with colors defaulting to identity (`pident`):

```{r part2-1}
# Show sequence alignment relationships
p3 <- ggchord(
  seq_data = seq_data_example,
  ribbon_data = ribbon_data_example
) +
  geom_seq() +
  geom_ribbon()
print(p3)
```

#### Customizing Ribbon Colors

Adjust ribbon color schemes via `ribbon_color_scheme` in `geom_ribbon()`:

```{r part2-2}
# Color by query sequence
p4 <- ggchord(
  seq_data = seq_data_example,
  ribbon_data = ribbon_data_example
) +
  geom_seq() +
  geom_ribbon(ribbon_color_scheme = "query")  # Ribbon color matches query sequence
print(p4)

# Single color
p5 <- ggchord(
  seq_data = seq_data_example,
  ribbon_data = ribbon_data_example
) +
  geom_seq() +
  geom_ribbon(
    ribbon_color_scheme = "single",  # Uniform color
    ribbon_colors = "orange"         # Custom color
  )
print(p5)
```

#### Impact of Sequence Layout on Ribbons

Ribbons automatically adapt to sequence orientation, curvature, and spacing:

```{r part2-4}
p6 <- ggchord(
  seq_data = seq_data_example,
  ribbon_data = ribbon_data_example
) +
  geom_seq(
    seq_orientation = c(1, -1, 1, -1),  # Sequence orientation
    seq_curvature = c(0, 2, -2, 6),     # Curvature
    seq_gap = c(0.1, 0.05, 0.09, 0.05), # Inter-sequence gaps
    seq_radius = c(1, 5, 1, 1)          # Sequence radii
  ) +
  geom_ribbon()
print(p6)
```

### 4.3 Adding Gene Annotation Information

Gene annotation data (`gene_data`) marks gene positions on sequences, with colors defaulting to strand direction (`strand`):

```{r part3-1}
# Show gene annotations
p7 <- ggchord(
  seq_data = seq_data_example,
  gene_data = gene_data_example
) +
  geom_seq() +
  geom_gene()
print(p7)
```

#### Customizing Gene Arrow Styles

Adjust gene colors via `gene_color_scheme` or display gene labels in `geom_gene()`:

```{r part3-2}
# Color by gene annotation category
p8 <- ggchord(
  seq_data = seq_data_example,
  gene_data = gene_data_example
) +
  geom_seq() +
  geom_gene(gene_color_scheme = "manual")  # Color by gene annotation (anno)
print(p8)

# Display gene labels and adjust styles
p9 <- ggchord(
  seq_data = seq_data_example,
  gene_data = gene_data_example,
  panel_margin = list(l = 0.2)  # Left margin
) +
  geom_seq() +
  geom_gene(
    gene_label_show = TRUE,          # Show labels
    gene_label_rotation = 45,        # Label rotation angle
    gene_label_radial_offset = 0.1   # Radial offset of labels
  )
print(p9)
```

## 5. Comprehensive Example

Using sequence, alignment, and gene data simultaneously with fine parameter adjustments, with each layer responsible for its own parameters:

```{r part4-2}
p10 <- ggchord(
  seq_data = seq_data_example,
  ribbon_data = ribbon_data_example,
  gene_data = gene_data_example,
  title = "Multiple Sequence Alignment Chord Diagram (with Gene Annotations)",  # Title
  rotation = 45  # Plot rotation angle
) +
  geom_seq(
    seq_gap = 0.03,                  # Inter-sequence gaps
    seq_radius = c(3, 2, 2, 1),      # Sequence radii
    seq_orientation = c(-1, -1, -1, 1),
    seq_curvature = c(0, 1, -1, 1.5)
  ) +
  geom_ribbon(
    ribbon_gap = 0.1,                # Ribbon gap
    ribbon_color_scheme = "pident"   # Color by identity
  ) +
  geom_gene(
    gene_offset = list(              # Gene arrow offset (strand-specific)
      c("+" = 0.2, "-" = -0.2),
      c("+" = 0.2, "-" = -0.2),
      c("+" = 0.2, "-" = 0),
      c("+" = 0.2, "-" = 0.1)
    ),
    gene_label_rotation = list(      # Gene label rotation (strand-specific)
      c("+" = 45, "-" = -45),
      c("+" = 0.2, "-" = -0.2),
      c("+" = 0.2, "-" = 0),
      c("+" = 0.2, "-" = 0.1)
    ),
    gene_width = 0.08,               # Gene arrow width
    gene_label_show = TRUE           # Show gene labels
  ) +
  geom_axis(
    axis_label_orientation = c(0, 45, 80, 130),  # Tick label angles
    axis_tick_major_length = 0.03,
    axis_label_size = 2
  )
print(p10)
```

## 6. Notes

- Layout is computed lazily when the plot is printed; all `geom_*` parameters are collected and applied at render time.
- The ribbon fill scale is independent from the gene fill scale, so both can be colored simultaneously.
- For parameter details, see the layer documentation: `?geom_seq`, `?geom_ribbon`, `?geom_gene`, and `?geom_axis`.
