---
title: "Plotting Multiple Sequence Alignment Chord Diagrams with ggchord"
author: "DangJem"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggchord Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 7
)
```

## 1. Introduction

`ggchord` is a ggplot2-based R package for plotting multiple sequence alignment chord diagrams, supporting visualization of alignment relationships between sequences and gene annotation information. This tutorial demonstrates its core functions and parameter usage through examples.

## 2. Installation and Loading

First, install and load the `ggchord` package (if not yet published on CRAN, install from local or GitHub):

```{r install, eval=FALSE}
# Install from local
devtools::install_local("path/to/ggchord_0.2.0.tar.gz")

# Load the package
library(ggchord)
library(dplyr)  # For data processing
```

```{r load, echo=FALSE}
library(ggchord)
library(dplyr)
```

## 3. Data Preparation

`ggchord` requires three types of data (sequence data is mandatory):  
- Sequence data (`seq_data`): Contains sequence IDs and lengths  
- Alignment data (`ribbon_data`): BLAST alignment results for drawing inter-sequence ribbons  
- Gene annotation data (`gene_data`): Contains gene positions and annotation information  

This tutorial uses built-in example data (added via `usethis::use_data()`):

```{r load-data}
# Load built-in example data
data(seq_data_example)    # Sequence data
data(ribbon_data_example) # BLAST alignment data
data(gene_data_example)   # Gene annotation data (short genes filtered out)

# Inspect data structure
head(seq_data_example)
head(ribbon_data_example)
head(gene_data_example)
```

## 4. Basic Usage

### 4.1 Using Only Sequence Data

Sequence data is fundamental for drawing chord diagrams, requiring at minimum `seq_id` (sequence ID) and `length` (sequence length):

```{r part1-1}
# Basic chord diagram: sequences only
p1 <- ggchord(
  seq_data = seq_data_example
)
print(p1)
```

#### Adjusting Sequence Parameters

Customize sequence order, orientation, curvature, and colors via parameters:

```{r part1-2}
p2 <- ggchord(
  seq_data = seq_data_example,
  seq_order = c("MT118296.1", "OR222515.1", "MT108731.1", "OQ646790.1"),  # Sequence order
  seq_orientation = c(1, -1, 1, -1),  # Direction (1 = forward, -1 = reverse)
  seq_curvature = c(0, 2, -2, 6),     # Curvature (0 = straight line, 1 = standard arc)
  seq_colors = c("steelblue", "orange", "pink", "yellow")  # Sequence colors
)
print(p2)
```

### 4.2 Adding Sequence Alignment Data

Alignment data (`ribbon_data`) visualizes sequence similarity, with colors defaulting to identity (`pident`):

```{r part2-1}
# Show sequence alignment relationships
p3 <- ggchord(
  seq_data = seq_data_example,
  ribbon_data = ribbon_data_example
)
print(p3)
```

#### Customizing Ribbon Colors

Adjust ribbon color schemes via `ribbon_color_scheme`:

```{r part2-2}
# Color by query sequence
p4 <- ggchord(
  seq_data = seq_data_example,
  ribbon_data = ribbon_data_example,
  ribbon_color_scheme = "query"  # Ribbon color matches query sequence
)
print(p4)

# Single color
p5 <- ggchord(
  seq_data = seq_data_example,
  ribbon_data = ribbon_data_example,
  ribbon_color_scheme = "single",  # Uniform color
  ribbon_colors = "orange"         # Custom color
)
print(p5)
```

#### Impact of Sequence Layout on Ribbons

Ribbons automatically adapt to sequence orientation, curvature, and spacing:

```{r part2-4}
p6 <- ggchord(
  seq_data = seq_data_example,
  ribbon_data = ribbon_data_example,
  seq_orientation = c(1, -1, 1, -1),  # Sequence orientation
  seq_curvature = c(0, 2, -2, 6),     # Curvature
  seq_gap = c(0.1, 0.05, 0.09, 0.05), # Inter-sequence gaps
  seq_radius = c(1, 5, 1, 1)          # Sequence radii
)
print(p6)
```

### 4.3 Adding Gene Annotation Information

Gene annotation data (`gene_data`) marks gene positions on sequences, with colors defaulting to strand direction (`strand`):

```{r part3-1}
# Show gene annotations
p7 <- ggchord(
  seq_data = seq_data_example,
  gene_data = gene_data_example
)
print(p7)
```

#### Customizing Gene Arrow Styles

Adjust gene colors via `gene_color_scheme` or display gene labels:

```{r part3-2}
# Color by gene annotation category
p8 <- ggchord(
  seq_data = seq_data_example,
  gene_data = gene_data_example,
  gene_color_scheme = "manual"  # Color by gene annotation (anno)
)
print(p8)

# Display gene labels and adjust styles
p9 <- ggchord(
  seq_data = seq_data_example,
  gene_data = gene_data_example,
  gene_label_show = TRUE,        # Show labels
  gene_label_rotation = 45,      # Label rotation angle
  gene_label_radial_offset = 0.1, # Radial offset of labels
  panel_margin = list(l = 0.2)    # Left margin
)
print(p9)
```

## 5. Comprehensive Example

Using sequence, alignment, and gene data simultaneously with fine parameter adjustments:

```{r part4-2}
p10 <- ggchord(
  seq_data = seq_data_example,
  ribbon_data = ribbon_data_example,
  gene_data = gene_data_example,
  title = "Multiple Sequence Alignment Chord Diagram (with Gene Annotations)",  # Title
  
  # Sequence parameters
  seq_gap = 0.03,
  seq_radius = c(3, 2, 2, 1),
  seq_orientation = c(-1, -1, -1, 1),
  seq_curvature = c(0, 1, -1, 1.5),
  
  # Gene parameters
  gene_offset = list(             # Gene arrow offset (strand-specific)
    c("+" = 0.2, "-" = -0.2),
    c("+" = 0.2, "-" = -0.2),
    c("+" = 0.2, "-" = 0),
    c("+" = 0.2, "-" = 0.1)
  ),
  gene_label_rotation = list(     # Gene label rotation (strand-specific)
    c("+" = 45, "-" = -45),
    c("+" = 0.2, "-" = -0.2),
    c("+" = 0.2, "-" = 0),
    c("+" = 0.2, "-" = 0.1)
  ),
  gene_width = 0.08,              # Gene arrow width
  gene_label_show = TRUE,         # Show gene labels
  
  # Ribbon parameters
  ribbon_gap = 0.1,
  ribbon_color_scheme = "pident", # Color by identity
  
  # Axis parameters
  axis_label_orientation = c(0, 45, 80, 130), # Tick label angles
  axis_tick_major_length = 0.03,
  axis_label_size = 2,
  
  # Overall style
  rotation = 45,  # Plot rotation angle
  show_axis = TRUE,
  panel_margin = list(t = 0.1)
)
print(p10)
```

## 6. Notes

- `ggchord` is currently in early development; some parameter combinations may cause graphical distortions, which will be optimized in future versions.
- Display effects of gene labels and ribbons may require multiple parameter adjustments for optimal results.
- For more parameter details, refer to the function documentation: `?ggchord`.
